---
title: "Petty Crime in Seattle Analysis"
author: "Ranch"
date: today
format:
  html:
    embed-resources: true
    message: false
    warning: false
    
---
## Loading the data
```{r}
echo = TRUE
results= "hide"
message = FALSE
warning = FALSE

library("tidyverse")

crime_data <- read.csv("SPD_Crime_Data__2008-Present_20250228.csv", stringsAsFactors = FALSE)

# displaying the crime data
crime_data |>
  glimpse()
```
## Filter data to violent crime
```{r}
violent_crime_data <- crime_data |>
  filter(Crime.Against.Category == "PERSON")
violent_crime_data
```



## Filter the data to only petty crime
From 1, 175, 457 rows, it became 995, 494 rows
```{r}
echo = TRUE
results='hide'
message = FALSE
warning = FALSE
petty_crime_data <- crime_data |>
  filter(Crime.Against.Category == "PROPERTY" | Crime.Against.Category == "SOCIETY")

petty_crime_data |>
  glimpse()
```
## Total petty crimes per month
```{r}
echo = FALSE
message = FALSE
warning = FALSE

library(plotly)

# convert to proper date time format 
petty_crime_data <- petty_crime_data |>
  mutate(Offense.Start.DateTime = mdy_hms(Offense.Start.DateTime, tz = "America/Los_Angeles"))

# Then: plot by month
plot <- petty_crime_data |>
  mutate(month = month(Offense.Start.DateTime, label = TRUE)) |>
  group_by(month) |>
  summarize(total_crimes = n()) |>
  ggplot(aes(x = month, y = total_crimes)) +
  geom_col(fill = "steelblue") +
  labs(
    title = "Petty Crimes in Seattle by Month",
    x = "Month",
    y = "Number of Petty Crimes"
  ) +
  theme_minimal()

str(petty_crime_data$Offense.Start.DateTime)

ggplotly(plot)
```

## Total petty crimes per year
```{r}
echo = FALSE
message = FALSE
warning = FALSE

# First, extract the year from Offense.Start.DateTime
petty_crime_data <- petty_crime_data %>%
  mutate(year = year(Offense.Start.DateTime))

# Group and count petty crimes per year
yearly_crimes <- petty_crime_data %>%
  group_by(year) %>%
  summarize(total_crimes = n())

# Plot
plot <- yearly_crimes |>
  ggplot(aes(x = year, y = total_crimes)) +
  geom_line(color = "darkgreen", linewidth = 1) +
  geom_point(color = "black", linewidth = 2) +
  scale_x_continuous(limits = c(2008,2024)) +
  labs(
    title = "Highest Petty Crime Count is 2020",
    x = "Year",
    y = "Number of Petty Crimes"
  ) +
  theme_minimal()
ggplotly(plot)
```

## Analyzing Petty Crimes by Category
We can break down these crimes by categories to observe if certain types of petty crimes are more prevalent
```{r}
echo = FALSE
message = FALSE
warning = FALSE

# Group by 'Crime.Against.Category' and 'month'
category_trends <- petty_crime_data %>%
  mutate(month = month(Offense.Start.DateTime, label = TRUE)) %>%
  group_by(month, Crime.Against.Category) %>%
  summarize(total_crimes = n())

# Plot crimes by category
plot <- category_trends |>
  ggplot(aes(x = month, y = total_crimes, fill = Crime.Against.Category)) +
  geom_col(position = "dodge") +
  labs(
    title = "There are more Property Crimes than Society Crimes in Seattle",
    subtitle = "Petty Crimes by Category in Seattle by Month",
    x = "Month",
    y = "Number of Crimes"
  ) +
  theme_minimal()
ggplotly(plot)
```

## Analyzing specific crime types and their frequency
We could go deeper and look at the specific crime types and how frequent they are
```{r}
echo = FALSE
message = FALSE
warning = FALSE

# Group by 'Offense.Parent.Group' and 'month'
crime_type_trends <- petty_crime_data %>%
  mutate(month = month(Offense.Start.DateTime, label = TRUE)) %>%
  group_by(month, Offense.Parent.Group) %>%
  summarize(total_crimes = n())

# Plot the crime types
plot <- crime_type_trends |>
  ggplot(aes(x = month, y = total_crimes, fill = Offense.Parent.Group)) +
  geom_col(position = "dodge") +
  labs(
    title = "Larcency Theft is the most Frequent Type of Crime",
    x = "Month",
    y = "Number of Crimes"
  ) +
  theme_minimal()
ggplotly(plot)
```

# Time of Day Analysis
We could look at how petty crimes are distributed throughout the day
```{r}
echo = FALSE
message = FALSE
warning = FALSE

# Extract hour of the day directly — no need to re-parse it
petty_crime_data <- petty_crime_data %>%
  mutate(hour = hour(Offense.Start.DateTime))

# Group by hour and plot
hourly_crimes <- petty_crime_data %>%
  group_by(hour) %>%
  summarize(total_crimes = n())

plot <- hourly_crimes |>
  ggplot(aes(x = hour, y = total_crimes)) +
  geom_line(color = "steelblue") +
  labs(
    title = "Petty Crimes often occur at midnight",
    x = "Hour of Day",
    y = "Number of Crimes"
  ) +
  theme_minimal()
ggplotly(plot)
```

## Heatmap of Violent Crime Data
```{r}
echo = FALSE
message = FALSE
warning = FALSE
library(plotly)
library(hexbin)

violent_crime_data <- violent_crime_data |>
  filter(Longitude > -125 & Longitude < -120,
         Latitude > 47 & Latitude < 48)

plot <- violent_crime_data |>
  ggplot(aes(x = Longitude, y = Latitude)) +
  geom_hex(bins = 50) +
  scale_fill_viridis_c(option = "inferno") +
  labs(
    title = "Violent Crime in Seattle from 2008 to Present",
    x = "Longitude",
    y = "Latitude",
    caption = "Data obtained from: https://data.seattle.gov/Public-Safety/SPD-Crime-Data-2008-Present/tazs-3rd5/about_data"
  ) + 
  theme_minimal()

ggplotly(plot)
```


## Heatmap of Petty Crime Data
```{r}
echo = FALSE
message = FALSE
warning = FALSE
library(hexbin)

petty_crime_data <- petty_crime_data |>
  filter(Longitude > -125 & Longitude < -120,
         Latitude > 47 & Latitude < 48)

plot <- petty_crime_data |>
  ggplot(aes(x = Longitude, y = Latitude)) +
  geom_hex(bins = 50) +
  scale_fill_viridis_c(option = "inferno") +
  labs(
    title = "Petty Crime in Seattle from 2008 to Present",
    x = "Longitude",
    y = "Latitude",
    caption = "Data obtained from: https://data.seattle.gov/Public-Safety/SPD-Crime-Data-2008-Present/tazs-3rd5/about_data"
  ) + 
  theme_minimal()

ggplotly(plot)
```

```{r}
library(ggplot2)
library(plotly)
library(dplyr)

# Add a hover text field with MCPP and coordinates
petty_crime_data <- petty_crime_data %>%
  mutate(hover_text = paste0(
    "Neighborhood: ", MCPP, "<br>",
    "Longitude: ", round(Longitude, 4), "<br>",
    "Latitude: ", round(Latitude, 4)
  ))

# Plot with geom_hex and pass hover text
plot <- ggplot(petty_crime_data, aes(x = Longitude, y = Latitude, text = hover_text)) +
  geom_hex(bins = 50, aes(fill = ..count..)) +
  scale_fill_viridis_c(option = "inferno", limits = c(0, 10000), name = "Crime Count") +
  labs(
    title = "Petty Crime in Seattle from 2008 to Present",
    x = "Longitude",
    y = "Latitude"
  ) +
  theme_minimal()

ggplotly(plot, tooltip = "text")
```


## Determing petty crime count on every neighborhood and by block address
```{r}
echo = FALSE
message = FALSE
warning = FALSE

# Crime count by MCPP (neighborhood)
crime_by_neighborhood <- petty_crime_data |>
  group_by(MCPP) |>
  summarise(Crime_Count = n()) |>
  arrange(desc(Crime_Count))

# View top neighborhoods
crime_by_neighborhood |> 
  head(10)

# Summarize crime count by block address
crime_by_block <- petty_crime_data |>
  group_by(X100.Block.Address) |>
  summarise(Crime_Count = n()) |>
  arrange(desc(Crime_Count))

# View top block addresses with the highest crime count
crime_by_block |> 
  head(10)
```


## Key Findings:
<ul>
  <li>Highest Number of Petty Crimes happened in 2020</li>
  <li>Larcency Theft is the most frequent type of crime</li>
  <li>Peak Crime Time -> Midnight</li>
  <li>There are more property crimes that society crimes</li>
  <li>The heatmap is similar to the original</li>
  <li>The top neighborhoods and block addresses are also similar</li>
</ul>