---
title: "Final Project"
author: "Rafael Juancho San Jose"
date: "2025-12-10"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, include=FALSE}
# libraries
library(readxl)
library(dplyr)
library(stargazer)

# load the data
df <- read_excel("WHR25_Data_Figure_2.1v3 (1).xlsx")

# rename variables
df <- df %>%
  rename(
    country = "Country name",
    life_eval = "Life evaluation (3-year average)",
    log_gdp = "Explained by: Log GDP per capita",
    social_support = "Explained by: Social support",
    healthy_life = "Explained by: Healthy life expectancy",
    freedom = "Explained by: Freedom to make life choices",
    generosity = "Explained by: Generosity",
    corruption = "Explained by: Perceptions of corruption"
  )

# filter for years 2021 - 2024
df <- df %>%
  filter(Year >= 2021, Year <= 2024) %>%
  
  # select only variables used for analysis
  select(Year, country, life_eval, log_gdp, social_support,
         healthy_life, freedom, generosity, corruption)

# remove NA values
df <- na.omit(df)
```
# Introduction

Across the world, national wealth is commonly assumed to translate directly into a better life for
citizens. Yet recent findings from the World Happiness Report reveal a puzzling pattern.
**Countries with high economic prosperity do not always achieve correspondingly high levels of life evaluation**.
The United States, for example, consistently ranks among the richest nations in terms of GDP per capita
but lags behind several lower-income countries in life satisfaction. This discrepancy raises an important
question about what truly drives human well-being and whether economic growth alone is sufficient to
improve people's lived experiences.

To contribute to understanding this puzzle, this project analyzes data from the **World Happiness Report (2021-2024)**,
covering **566 country-year observations** across dozens of nations. The **unit of analysis** is the
**country-year**, and the outcome variable is **life evaluation**, a 0-10 measure of how individuals
assess their overall life quality. The main explanatory variable is **log GDP per capita**, representing
economic prosperity. Additional variables are **social support**, **healthy life expectancy**,
**freedom to make life choices**, **generosity**, **perceptions of corruption**. This leads to the central
research question: **How strongly is log GDP per capita associated with life evaluation across countries, and do social and institutional factors help explain cross national differences?**


# Theory and Hypothesis

Understanding why economic prosperity does not always translate into higher life evaluation has been a
central question in the study of well-being. While GDP per capita is often used as a proxy for national
progress, prior research shows that income alone may be insufficient to explain how people assess their
overall life satisfaction. This project seeks to explain **why countries with similar levels of economic development differ in reported well-being**,
and whether social and institutional conditions account for part of this variation.

A large body of research in economics and psychology highlights the limits of income as a driver of
subjective well-being. The **Easterlin Paradox** argues that increases in national income do not
continuously raise happiness once basic needs are met, suggesting diminishing returns to wealth
(Easterlin 1974). Other studies emphasize the role of **social supportc**, **trust in institutions**, and **health** in shaping individuals' life satisfaction **(Helliwell & Putnam 2004; Deaton 2008; Helliwell 2003)**. 
Cross-national analyses consistently find that people report higher well- being when they feel supported by communities, perceive governments as fair and trustworthy, enjoy strong physical health and longevity
**(Diener & Biswas-Diener 2002; Inglehart et al. 2008)**. Together, these findings imply that well-being is multidimensional and shaped by more than economic resources alone.

Based on this literature, the causal mechanism is that **economic prosperity (X)** influences **life evaluation (Y)**
partially through its effect on social and institutional factors. Income may improve well-being by
enabling access to healthcare, improving living standards, and supporting stable social environments.
However, when social support is weak, corruption is high, or individuals lack freedom in making life
choices, these institutional shortcomings may limit the impact of GDP on subjective well-being. Thus,
countries with similar economic output may yield different well-being outcomes depending on social
structures and governance quality.

From this theoretical foundation, the following hypotheses are developed:

- $H_0$ (Null Hypothesis): Log GDP per capita has no association with life evaluation once social support,
health, freedom, and corruption are accounted for.
$$
\beta_{log~GDP} = 0
$$
- $H_1$ (Alternative Hypothesis): Log GDP per capita is positively associated with life evaluation even
after accounting for social support, health, freedom, and corruption.
$$
\beta_{log~GDP} > 0
$$
These hypotheses reflect the expectation supported by previous research that economic prosperity improves
well-being, but its impact is only partial, with social and institutional factors also playing meaningful
roles.


# Data and Visual Analysis
## Data

This project uses data from **World Happiness Report** covering years **2021-2024**. After filtering to
include only these years and removing incomplete cases, the final dataset contains **566 country-year observations**.
The **unit of analysis** is the country-year, meaning each row represents a specific country observed in
a specific year.

The dataset is **observational**, not experimental because no treatment was randomly assigned, causal
inference is limited. Instead the analysis focuses on identifying associations between economic, social,
and institutional factors and the outcome variable, **Life Evaluation**. Potential confounding is likely,
especially because variables such as social support, perceptions of corruption, and health are correlated
with each other and with GDP. No variable transformations were necessary beyond renaming, since all 
variables are already scaled on comparable 0-10 Cantril ladder units.


```{r, results='asis', echo=FALSE, message=FALSE, warning=FALSE, fig.align='center'}
dat_des <- df[,c("life_eval", "log_gdp", "social_support", "healthy_life", "freedom", "generosity",
                   "corruption")]

stargazer(as.data.frame(dat_des), # input the object "as.data.frame" for descriptives
          summary.stat = c("mean","sd","min","median","max"),
          type="latex",
          header = FALSE,
          title = "Descriptive statistics",
          digits = 2, 
          float = FALSE)

# Display a correlation matrix

# create an object with the correlation matrix

dat_cor <- cor(dat_des)

# Now we need to format the correlation matrix, start rounding the digits

dat_cor <- round(dat_cor, digits = 2)

# display the correlation matrix in a formmatted table using stargazer

stargazer(dat_cor, 
          type="latex",
          title = "Correlation matrix",
          column.sep.width = "1pt",
          header = FALSE, 
          digits = 2,
          float = FALSE)
```


The table shows considerable variation across the dataset. **Life Evaluation** averages 5.55, with a
minimum of 1.36 and maximum of 7.82, indicating substantial differences in perceived well-being across
countries. **Log GDP per capita** is moderately concentrated around 1.38, while **social support** and
**healthy life expectancy** show smaller ranges, suggesting these factors vary less between countries.
**Freedom** and **perceptions of corruption** show moderate dispersion, and **generosity** is generally
low with minimal variation.

The correlation matrix shows the pairwise linear relationships among key variables. Life evaluation is
most strongly associated with Log GDP (r = 0.76) and Social Support (r = 0.74), followed by Healthy Life
(r = 0.67) and Freedom to make life choices (r = 0.56), indicating that higher economic prosperity,
stronger social networks, better health, and greater freedom tend to correspond to higher life
satisfaction across countries. In contrast, Generosity shows almost no linear association with Life 
Evaluation (r = 0.07), suggesting it is not a meaningful predictor in this context. Among the explanatory variables, some moderate correlations exist. For example, Log GDP and Healthy Life are correlated at
0.71. Though no pair of variables exceeds levels, they are typically considered problematic for multicollinearity. Overall, the matrix highlights which factors are most strongly related to Life
Evaluation and supports the selection of Log GDP, Social Support, Healthy Life, Freedom, and Corruption
as primary predictors in the regression analysis.


## Exploratory Data Analysis

```{r, warning = FALSE, echo = FALSE, message = FALSE}
# install.packages("ggcorrplot")

library(ggcorrplot)

cor_mat <- round(cor(df[, c("life_eval","log_gdp","social_support","healthy_life",
                            "freedom","generosity","corruption")]), 2)

ggcorrplot(cor_mat, 
           lab = TRUE,
           title = "Correlation Heatmap of Key Variables",
           ggtheme = ggplot2::theme_minimal())
```

To explore the relationships among the key variables in the World Happiness dataset, I began by examining
a correlation heatmap. This visualization provides a compact view of linear associations and reveals 
clear clusters among variables related to well-being. Life evaluation is strongly correlated with Log GDP per capita, social support, and healthy life expectancy, suggesting that economic and social conditions 
are major contributors to national well-being. Freedom shows a moderate positive relationship, while 
generosity and perceptions of corruption display weaker associations. These patterns indicate that GDP, 
social support, and health are the most promising predictors to include in the regression model, while 
generosity is unlikely to be influential.

```{r, warning = FALSE, echo = FALSE, message = FALSE}
library(ggplot2)

ggplot(df, aes(x = log_gdp, y = life_eval)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", se = TRUE) +
  labs(title = "GDP per Capita vs Life Evaluation",
       x = "Log GDP per Capita",
       y = "Life Evaluation") +
  theme_minimal()
```
The scatterplot shows a strong positive linear relationship between log GDP per capita and life 
evaluation. Countries with higher economic prosperity correlate with higher levels of subjective 
well-being. The fitted regression line slopes upward, and the data points cluster tightly around it, 
suggesting that GDP is one of the strongest predictors of life evaluation. This visual reinforces the 
idea that economic resources provide important foundations for well-being but may not fully explain 
cross-national differences.

```{r, warning = FALSE, echo = FALSE, message = FALSE}
ggplot(df, aes(x = social_support, y = life_eval)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", se = TRUE) +
  labs(title = "Social Support vs Life Evaluation",
       x = "Social Support",
       y = "Life Evaluation") +
  theme_minimal()
```
The plot demonstrates a moderate positive relationship between perceived freedom and life evaluation. 
While the upward slope is clear, the points are more dispersed compared to previous variables. This 
suggests that freedom meaningfully contributes to well-being, but its relationship is weaker and more 
context-dependent. Still, the visual indicates that autonomy and perceived control over life decisions 
are relevant predictors.

```{r, warning = FALSE, echo = FALSE, message = FALSE}
ggplot(df, aes(x = corruption, y = life_eval)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", se = TRUE) +
  labs(title = "Perceptions of Corruption vs Life Evaluation",
       x = "Perceptions of Corruption (Higher = Less Corruption)",
       y = "Life Evaluation") +
  theme_minimal()
```
This scatterplot shows a positive slope, indicating that lower perceived corruption is associated with 
higher life evaluation. The relationship is weaker and more scattered, but the general pattern is upward. 
This suggests that governance quality and institutional trust contribute to subjective well-being, 
although the effect is smaller relative to GDP, social support, or health.


# Statistical Analysis

To evaluate the relationship between GDP per capita and life evaluation, I estimate a multiple linear 
regression model that includes log GDP per capita as the main predictor and several theoretically 
grounded confounders: social support, healthy life expectancy, freedom to make life choices, and 
perceptions of corruption. These variables are included because they influence both national economic 
conditions and subjective well-being, and therefore satisfy the structure of potential confounders 
(log_gdp ← Z → life_eval).

This specification allows us to isolate the association between GDP per capita and life evaluation while 
accounting for key social and institutional factors. The resulting model provides evidence on whether GDP 
per capita remains a significant predictor of life evaluation after adjusting for these confounding 
influences.
```{r, results='asis', echo=FALSE, message=FALSE, warning=FALSE, fig.align='center'}
fit <- lm(life_eval ~ log_gdp + social_support + healthy_life + freedom + corruption, data = df)

stargazer(fit,
          type = "latex",
          style = "ajps",
          column.sep.width = "1pt",
          title = "Regression analysis of Log GDP on Life Evaluation",
          header = FALSE,
          omit.stat=c("f", "ser"))
```


# Results Discussion

The regression results provide strong evidence against the null hypothesis, indicating that GDP per 
capita is positively associated with life evaluation even after accounting for major confounders. Log GDP 
remains highly significant in all specifications, suggesting a robust relationship that does not 
disappear once social support, healthy life expectancy, freedom, and corruption are included. This 
consistency strengthens confidence that GDP captures a meaningful dimension of subjective well-being 
across countries.


At the same time, the results show that several non-economic factors contribute comparably or even more 
strongly to life evaluation. Social support, freedom, health, and perceived corruption all remain 
significant predictors in the fully adjusted model, underscoring that subjective well-being is 
multidimensional rather than purely economic. The stability of these findings across specifications 
indicates that the key associations are not driven by model choice or omitted-variable bias within the 
available data.

While the results are strong, they rely on cross-sectional comparisons between countries rather than 
changes within countries over time. Future work could use panel or longitudinal data to examine whether 
increases in GDP or improvements in social conditions lead to higher life evaluation within the same 
country. Additional variables—such as more detailed health measures, cultural factors, or institutional 
quality indicators—could also help clarify the mechanisms behind the observed relationships.


# Conclusion

This project examined the relationship between GDP per capita and life evaluation using World Happiness 
Report data from 2021–2024 across 566 country-year observations. The research question asked whether 
higher national income corresponds to higher subjective well-being, and the analysis incorporated major 
confounders—social support, freedom, healthy life expectancy, and perceptions of corruption—to more 
accurately estimate this relationship.

The findings indicate strong support for the hypothesis: GDP per capita is positively and significantly 
associated with life evaluation. However, the results also reveal that several social and institutional 
factors play substantial and independent roles. Social support, health, freedom, and corruption are all 
significant predictors, suggesting that economic conditions alone cannot fully explain cross-country 
variation in well-being.

Overall, the evidence supports the view that subjective well-being is shaped by a combination of economic
prosperity and broader social and institutional environments. While GDP matters, countries with strong 
social support systems, healthier populations, greater freedoms, and lower corruption tend to exhibit 
higher life evaluations regardless of income levels. These findings reinforce the importance of 
multidimensional policy approaches to improving national well-being.